---
title: "Muscle Proteomic Response to Blood Flow Restriction versus High-Load Resistance Training"
subtitle: "Preliminary Analysis Report — BFR vs. HLRT"
author: "NAMPL"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
geometry: margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{grffile}
  - \usepackage{fancyhdr}
  - \usepackage{booktabs}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[R]{\small BFR Proteomics — Preliminary Report}
  - \fancyfoot[C]{\thepage}
  - \renewcommand{\headrulewidth}{0.4pt}
  - \renewcommand{\maketitle}{}
---

<!--
HOW TO USE THIS FILE
====================
1. Save this in your project root (same level as 01_normalization/, 03_analysis/,
   04_figures/), or adjust `base_dir` in the setup chunk below.
2. If you've never knit to PDF before, you'll need a LaTeX distribution:
     install.packages("tinytex")
     tinytex::install_tinytex()
3. In RStudio: use the dropdown arrow next to "Knit" (not just the button
   itself) and choose "Knit to PDF" -- RStudio remembers the last format
   you picked per-document, so if this file (or one you copied it from)
   was ever knit to HTML before, the button can default to HTML even
   though the YAML above says pdf_document. Alternatively, from the
   console: rmarkdown::render("00_project_report.Rmd", output_format = "pdf_document")
   which ignores the remembered button state entirely.
4. Search for "EDIT ME" throughout this file -- those are the spots that need
   your study-specific narrative (PI name, sample sizes, biological
   interpretation, etc.). Everything else pulls automatically from your
   pipeline's saved CSVs/figures, so re-knitting after a pipeline re-run
   updates the numbers/plots without you touching this file.
5. Any figure/table that isn't found on disk yet will show as a placeholder
   note instead of breaking the render, so you can knit this at any point
   in the pipeline, not just at the very end.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(dplyr, readr, knitr, kableExtra)

# --- Project root -------------------------------------------------------
# Robust resolution + diagnostics. If figures/tables show "not yet
# available" even though the files exist, it means base_dir below isn't
# pointing at your actual project folder -- check the printed values
# first before looking anywhere else.
base_dir <- getwd()

# rprojroot::find_rstudio_root_file() requires an actual .Rproj file to
# exist; if getwd() doesn't already look like the project root (no
# 01_normalization/ folder under it), try rprojroot as a fallback before
# giving up.
if (!dir.exists(file.path(base_dir, "01_normalization"))) {
  base_dir <- tryCatch(rprojroot::find_rstudio_root_file(), error = function(e) getwd())
}

cat("Resolved base_dir:", base_dir, "\n")
cat("  01_normalization/ found here:", dir.exists(file.path(base_dir, "01_normalization")), "\n")
cat("  04_figures/       found here:", dir.exists(file.path(base_dir, "04_figures")), "\n")

if (!dir.exists(file.path(base_dir, "01_normalization"))) {
  warning(
    "base_dir does not contain your pipeline folders (01_normalization/, ",
    "04_figures/, etc.) -- every file path below will resolve incorrectly. ",
    "Hardcode the correct path manually, e.g.:\n",
    '  base_dir <- "/Users/yourname/Desktop/YourProjectFolder"\n',
    "immediately after this warning, then re-knit."
  )
}

knitr::opts_knit$set(root.dir = base_dir)

# --- Paths to existing pipeline outputs ----------------------------------
# All relative to base_dir; nothing here re-runs analysis, it only reads
# files your normalization / GSEA / figure scripts already wrote.
p <- list(
  # Normalization: data
  sample_summary_raw = "01_normalization/c_data/00_raw_sample_summary.csv",
  protein_report      = "01_normalization/c_data/11_protein_report_BFR.csv",

  # Normalization: reports
  sample_report        = "01_normalization/b_reports/12_sample_report_BFR.csv",
  filtering_effects     = "01_normalization/b_reports/02_filtering_effects_BFR.csv",
  filtering_plot        = "01_normalization/b_reports/04_filtering_plot_BFR.pdf",
  contam_bm_plot        = "01_normalization/b_reports/01_blood_v_muscle_intensity_plot_raw.pdf",
  outlier_diag_plot     = "01_normalization/b_reports/06_outlier_diagnostic_plots_BFR.pdf",
  pca_raw               = "01_normalization/b_reports/09_pca_raw_BFR.pdf",
  pca_postnorm           = "01_normalization/b_reports/09_pca_postnorm_BFR.pdf",
  cv_violin_prenorm      = "01_normalization/b_reports/07_cv_violin_BFR_prenorm.pdf",
  cv_violin_norm         = "01_normalization/b_reports/07_cv_violin_BFR_normalized.pdf",

  # Analysis figures (fig_02)
  fig02_da    = "04_figures/fig_02/fig_02_A.png",
  fig02_gsea  = "04_figures/fig_02/fig_02_B.png",
  fig02_upset = "04_figures/fig_02/fig_02_C.png",
  fig02_cv    = "04_figures/fig_02/fig_02_D.png",

  # Analysis figures (fig_03)
  fig03_volcano = "04_figures/fig_03/panel_3A.png",

  # Analysis figures (protein-level intensity: PRE/POST by contrast)
  intensity_bfr_up      = "04_figures/fig_protein_intensity/intensity_Training_BFR_up.png",
  intensity_bfr_down    = "04_figures/fig_protein_intensity/intensity_Training_BFR_down.png",
  intensity_hlrt_up     = "04_figures/fig_protein_intensity/intensity_Training_HLRT_up.png",
  intensity_hlrt_down   = "04_figures/fig_protein_intensity/intensity_Training_HLRT_down.png",
  intensity_both_up     = "04_figures/fig_protein_intensity/intensity_both_concordant_up.png",
  intensity_both_down   = "04_figures/fig_protein_intensity/intensity_both_concordant_down.png",

  # Analysis figures (GSEA bubble plot)
  gsea_bubble_BFR = "04_figures/fig_gsea_bubble/gsea_bubble_BFR_Training_BFR_top30_percontrast.png",
  gsea_bubble_HLRT = "04_figures/fig_gsea_bubble/gsea_bubble_BFR_Training_HLRT_top30_percontrast.png",

  # Analysis figures (GSEA differential: Training_BFR vs Training_HLRT)
  gsea_diff_counts  = "04_figures/fig_gsea_differential/gsea_differential_counts.png",
  gsea_diff_scatter = "04_figures/fig_gsea_differential/gsea_differential_scatter.png"
)
p <- lapply(p, function(x) file.path(base_dir, x))

# --- Helpers: fail gracefully instead of breaking the knit ---------------

# Raw LaTeX (used below for [H] hard placement) doesn't get pandoc's
# automatic escaping of special characters the way markdown text does.
# Underscores especially -- e.g. "Training_BFR" in a caption -- are
# reserved math-mode characters in LaTeX and cause a "Missing $ inserted"
# error if passed through unescaped. grffile (in header-includes) handles
# underscores in \includegraphics filenames; this handles them (and other
# LaTeX special characters) in caption text, which grffile doesn't cover.
escape_latex <- function(x) {
  x <- gsub("\\", "\u0001BACKSLASH\u0001", x, fixed = TRUE)
  x <- gsub("&", "\\&", x, fixed = TRUE)
  x <- gsub("%", "\\%", x, fixed = TRUE)
  x <- gsub("$", "\\$", x, fixed = TRUE)
  x <- gsub("#", "\\#", x, fixed = TRUE)
  x <- gsub("_", "\\_", x, fixed = TRUE)
  x <- gsub("{", "\\{", x, fixed = TRUE)
  x <- gsub("}", "\\}", x, fixed = TRUE)
  x <- gsub("~", "\\textasciitilde{}", x, fixed = TRUE)
  x <- gsub("^", "\\textasciicircum{}", x, fixed = TRUE)
  x <- gsub("\u0001BACKSLASH\u0001", "\\textbackslash{}", x, fixed = TRUE)
  x
}

include_fig <- function(path, caption = "", width = "85%") {
  if (file.exists(path)) {
    # Raw LaTeX instead of markdown image syntax: [H] (hard placement,
    # via the float package loaded in header-includes) forces the figure
    # to render exactly here rather than letting LaTeX's automatic float
    # algorithm push it elsewhere in the document -- pandoc's markdown
    # image syntax has no placement control, which is what was causing
    # figures to drift away from their position in the .Rmd source.
    width_frac <- as.numeric(sub("%", "", width)) / 100
    cat(sprintf(
      "\\begin{figure}[H]\n\\centering\n\\includegraphics[width=%.2f\\linewidth]{%s}\n\\caption{%s}\n\\end{figure}\n\n",
      width_frac, path, escape_latex(caption)
    ))
  } else {
    cat(sprintf("*[Figure not yet available: `%s` — run the corresponding pipeline step to generate it.]*\n\n",
                basename(path)))
  }
}

include_table <- function(path, caption = "", max_rows = 20) {
  if (file.exists(path)) {
    df <- read.csv(path)
    if (nrow(df) > max_rows) {
      cat(sprintf("*(showing first %d of %d rows)*\n\n", max_rows, nrow(df)))
      df <- head(df, max_rows)
    }
    # HOLD_position (kableExtra) = hard [H] placement, same rationale as
    # include_fig above -- without this, tables float away from their
    # position in the source too.
    print(
      knitr::kable(df, caption = caption, row.names = FALSE, booktabs = TRUE) %>%
        kableExtra::kable_styling(latex_options = c("HOLD_position"))
    )
  } else {
    cat(sprintf("*[Table not yet available: `%s`]*\n\n", basename(path)))
  }
}
```

\begin{titlepage}
\centering
\vspace*{2cm}

{\LARGE \bfseries Muscle Proteomic Response to Blood Flow Restriction versus High-Load Resistance Training}

\vspace{0.5cm}
{\Large Preliminary Analysis Report}

\vspace{1.5cm}
{\large Prepared for: \textbf{Dr. Cleiton Libardi}}

\vspace{0.3cm}
{\large Prepared by: \textbf{Sam Norton and Tahran Gotla}}

\vspace{0.3cm}
{\large `r format(Sys.Date(), '%B %d, %Y')`}

\vspace{2cm}
\begin{minipage}{0.8\textwidth}
\small
\textit{This report summarizes data preprocessing, quality control, normalization,
differential abundance, and pathway enrichment analyses completed to date for the
BFR vs. HLRT muscle proteomics study. Results are preliminary and subject to
revision as the analysis pipeline continues.}
\end{minipage}

\vfill
\end{titlepage}

\newpage
\tableofcontents
\newpage

# Project Overview

An examination of the proteomic alterations induced by blood flow restriction
(BFR) vs. high-load resistance training (HLRT) in human vastus lateralis biopsy 
samples. N=33 participants' legs were randomly assigned to BFR or HLRT. Biopsies
were donated PRE and POST training.

**Proteomic analysis design at a glance:**

```{r contrast-table}
contrast_tbl <- tibble::tribble(
  ~Comparison,      ~Contrast,                    ~`Biological question`,
  "Baseline",       "BFR_PRE vs HLRT_PRE",        "Do the two arms differ before training begins?",
  "Training_BFR",   "BFR_POST vs BFR_PRE",        "How does the muscle proteome change with BFR training?",
  "Training_HLRT",  "HLRT_POST vs HLRT_PRE",      "How does the muscle proteome change with HLRT training?",
  "Post_training",  "BFR_POST vs HLRT_POST",      "Do the two arms differ after training?",
  "Interaction",    "(BFR_POST-BFR_PRE) vs (HLRT_POST-HLRT_PRE)", "Does the training response itself differ by modality?"
)
knitr::kable(contrast_tbl, booktabs = TRUE) %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position")) %>%
  kableExtra::column_spec(1, width = "1in") %>%
  kableExtra::column_spec(2, width = "1.9in") %>%
  kableExtra::column_spec(3, width = "3.1in")
```

```{r sample-summary-text, results='asis'}
include_table(p$sample_report,
             caption = "Final sample counts per group after filtering and outlier removal: paired (both PRE and POST present), PRE-only, and POST-only subjects, with subject and sample totals.")
```

\newpage

# Pipeline Overview

The analysis pipeline proceeds through the following stages, each detailed
in the sections below:

1. **Data preprocessing** — raw intensity/metadata loading, annotation cleanup, rerun consolidation
2. **Contaminant assessment** — tissue-specificity (HPA) and blood-protein flagging
3. **Filtering** — contaminant removal and per-group missingness filtering
4. **Outlier detection** — consensus-based sample QC
5. **Normalization** — cyclic loess normalization
6. **Differential abundance analysis** — limma, 5 contrasts
7. **Pathway / enrichment analysis** — GSEA + ORA across 7 databases, with redundancy filtering

\newpage

# 1. Data Preprocessing

Raw protein-group intensities and sample metadata were loaded and restructured
prior to any filtering:

- Protein annotation columns (UniProt ID, protein name, gene symbol,
  description) were parsed from the search-engine output, and multi-protein
  group matches were collapsed to their first (primary) match.
- cRAP (common Repository of Adventitious Proteins) contaminant tags were
  flagged directly from the raw protein group string, before any trimming
  could drop that information.
- Sample columns were cleaned to a consistent `S###` naming convention.
- **Rerun consolidation:** samples with a trailing "r"/"R" suffix (indicating
  a repeat MS injection) were identified and merged onto their base sample
  name, with a `rerun_level` flag retained in the metadata for traceability.
- Metadata and intensity columns were validated to be in 1:1 correspondence
  before proceeding (a hard `stopifnot` check — the pipeline halts rather
  than silently misaligning samples).

\newpage

# 2. Contaminant Assessment

Two independent contamination checks were applied to every protein, in
addition to a per-sample contamination index:

**A. Tissue specificity (Human Protein Atlas).** Each protein's UniProt ID
was checked against the HPA skeletal muscle gene list; proteins absent from
this reference were flagged as not muscle-specific.

**B. Blood protein blacklist.** A curated list of classic blood/plasma
contaminant gene families was applied — albumin & lipid transport proteins,
coagulation factors, complement system components, hemoglobins, acute-phase
proteins, and platelet/neutrophil markers — supplemented dynamically with
any immunoglobulin (IGH/IGK/IGL) or keratin (KRT) genes detected in this
specific dataset.

**C. Sample-level contamination index.** For every sample, the summed
intensity of blood markers (ALB, HBB, HBA1) vs. muscle markers (MB, CKM)
was expressed as a percentage of total intensity, giving a blood:muscle
(B:M) ratio per sample — a diagnostic for samples with unusually high blood
contamination (e.g. from incomplete blood clearance during biopsy).

```{r contam-bm-plot, results='asis'}
include_fig(p$contam_bm_plot, caption = "Blood vs. muscle marker proportion of total intensity, per sample.")
```

\newpage

# 3. Filtering

Proteins and samples were filtered in the following order:

1. **Zero $\rightarrow$ missing.** Raw zero-intensity values (an artifact of the
   search engine, not a true measurement of zero abundance) were converted
   to `NA`.
2. **HPA muscle filter.** Proteins not annotated as muscle-expressed in the
   Human Protein Atlas were removed.
3. **Blood blacklist filter.** Proteins matching the blood contaminant gene
   list (Section 2B) were removed.
4. **Missingness filter.** Proteins were required to be observed in at
   least 50% of samples *within* at least one group (BFR or HLRT), rather
   than 50% of the full cohort — this retains proteins that are
   consistently detected in one arm even if largely absent in the other,
   which a whole-cohort threshold would have discarded.

```{r filtering-plot, results='asis'}
include_fig(p$filtering_plot, caption = "Number of proteins retained at each filtering step.", width = "70%")
```

```{r filtering-table, results='asis'}
include_table(p$filtering_effects, caption = "Filtering effects: proteins before/after each step.")
```

\newpage

# 4. Outlier Detection

Samples were screened for exclusion using a **consensus of three
independent methods**, each flagging samples that are statistical outliers
relative to the rest of the cohort:

- **Missingness outliers:** samples with % missing proteins beyond
  1.5$\times$IQR above the 75th percentile.
- **PCA outliers:** samples with a Mahalanobis distance (computed on the
  first 3 principal components of log2 intensities) exceeding the
  99th-percentile $\chi^2$ threshold.
- **Intensity outliers (MAD):** samples whose median log2 intensity
  deviates from the cohort median by more than 3$\times$ the median absolute
  deviation (MAD).

A sample was excluded only if flagged by **at least 2 of the 3** methods,
to avoid removing samples based on a single noisy metric.

```{r outlier-plot, results='asis'}
include_fig(p$outlier_diag_plot, caption = "Outlier diagnostics: missingness, PCA (Mahalanobis), and MAD intensity.")
```

\newpage

# 5. Normalization

Following filtering and outlier removal, protein intensities were
normalized using **cyclic loess normalization**, which iteratively aligns
the intensity distribution of each sample against the others via local
regression (loess) on pairwise MA-plots. This corrects for
sample-to-sample technical variation (e.g. differences in total protein
loaded, digestion/labeling efficiency) without assuming the majority of
proteins are unchanged between the two study arms, which a simpler
median/quantile normalization would.

**Quality control before vs. after normalization:**

```{r pca-raw, results='asis'}
include_fig(p$pca_raw, caption = "PCA of raw (pre-normalization) intensities, colored by group.", width = "48%")
```

```{r pca-postnorm, results='asis'}
include_fig(p$pca_postnorm, caption = "PCA of normalized intensities, colored by group.", width = "48%")
```

```{r cv-prenorm, results='asis'}
include_fig(p$cv_violin_prenorm, caption = "Per-protein coefficient of variation (CV) by group, pre-normalization.", width = "60%")
```

```{r cv-norm, results='asis'}
include_fig(p$cv_violin_norm, caption = "Per-protein coefficient of variation (CV) by group, post-normalization.", width = "60%")
```

\newpage

# 6. Differential Abundance Analysis

Differential abundance was assessed with **limma**, fit across 5 contrasts
representing the study's 2 $\times$ 2 (arm $\times$ training) design plus the arm$\times$time
interaction (see the contrast table in Section "Project Overview"). For
each contrast, three complementary significance metrics were retained: the
raw p-value, the Benjamini-Hochberg FDR-adjusted p-value, and a composite
**Pi-value** (p^|log2FC|; Xiao et al., 2014)
used as the primary threshold (Pi-value < 0.05) for calling a protein
differentially abundant.

```{r fig02-da, results='asis'}
include_fig(p$fig02_da, caption = "Number of differentially abundant proteins (DAPs) per contrast, by significance method.", width = "60%")
```

```{r fig02-upset, results='asis'}
include_fig(p$fig02_upset, caption = "Overlap of differentially abundant proteins across contrasts (UpSet plot), colored by direction of change.")
```

\newpage

## Volcano Plots by Contrast

Per-protein volcano plots for each of the 5 contrasts, showing effect size
(log2 fold change) against significance (-log10 p-value). Proteins are
colored by direction of change, with the top proteins by fold-change
magnitude labeled in each panel and UP/DOWN counts annotated in the
corners.

```{r fig03-volcano, results='asis'}
include_fig(p$fig03_volcano, caption = "Volcano plots for all 5 contrasts.", width = "95%")
```

\newpage

## Protein-Level Intensity Plots

**Proteins concordantly significant in both contrasts.** The following
show proteins that were significantly up- or down-regulated with the
*same direction* in both Training_BFR and Training_HLRT (ranked so both
contrasts' p-values must be small, not just one), with all four
group $\times$ timepoint combinations shown per protein (BFR PRE/POST,
HLRT PRE/POST) to visualize the shared response pattern across both
training modalities directly.

```{r intensity-both-up, results='asis'}
include_fig(p$intensity_both_up, caption = "Top 20 up-regulated proteins, significant in both Training_BFR and Training_HLRT.", width = "95%")
```

```{r intensity-both-down, results='asis'}
include_fig(p$intensity_both_down, caption = "Top 20 down-regulated proteins, significant in both Training_BFR and Training_HLRT.", width = "95%")
```

\newpage

# 7. Pathway / Enrichment Analysis

For each of the 5 contrasts, two complementary enrichment approaches were
run:

**Gene Set Enrichment Analysis (GSEA)** — rank-based, using the limma
t-statistic as the ranking metric, tested against 7 pathway/gene-set
databases: GO Biological Process, GO Cellular Component, GO Molecular
Function (via `clusterProfiler::gseGO`), a GO Slim subset, KEGG, Reactome,
and MSigDB Hallmark gene sets.

**Over-Representation Analysis (ORA)** — tested the sets of significantly
up- and down-regulated proteins (Pi-value < 0.05) separately against GO
BP/CC/MF, using all detected proteins as the statistical background.
NOTE: no ORA results are included in this report. 

**Redundancy filtering.** Because gene set databases contain many
near-duplicate terms (e.g. highly overlapping GO terms), a pooled greedy
redundancy filter was applied within each contrast $\times$ database: significant
pathways were ranked by adjusted p-value, and a pathway was dropped if its
gene-set membership overlapped an already-retained, more-significant
pathway above a threshold (Jaccard index $\geq$ 0.5 **or** overlap coefficient
$\geq$ 0.5). This keeps the most significant representative of each cluster of
redundant terms rather than reporting near-duplicates as independent hits.

```{r fig02-gsea, results='asis'}
include_fig(p$fig02_gsea, caption = "Number of significant (non-redundant) enriched pathways per contrast, by database.", width = "60%")
```

## GSEA Bubble Plot

Significant Hallmark and GO Biological Process pathways for the
Training_BFR and Training_HLRT contrasts, restricted to redundancy-filter
representative pathways (Section 7's redundancy filtering, applied above
before ranking). Point size reflects statistical confidence
(-log~10~ FDR), fill color reflects normalized enrichment score (NES,
red = up in that contrast, blue = down), and outline indicates whether
that specific point clears the FDR < 0.05 threshold.

```{r gsea-bubble-BFR, results='asis'}
include_fig(p$gsea_bubble_BFR, caption = "GSEA bubble plot: Top 30 pathways significant in the Training_BFR contrast across 7 databases.", width = "90%")
```

```{r gsea-bubble-HLRT, results='asis'}
include_fig(p$gsea_bubble_HLRT, caption = "GSEA bubble plot: Top 30 pathways significant in the Training_HLRT contrast across 7 databases.", width = "90%")
```

\newpage

## Differential Pathway Analysis: BFR vs. HLRT Training Response

To assess whether the two training modalities engage the same or
different biological pathways, every redundancy-filtered pathway was
classified by its significance and direction (NES sign) in Training_BFR
vs. Training_HLRT: **BFR only** or **HLRT only** (significant in one
contrast but not the other), **Discordant** (significant in both, but
opposite direction), or **Concordant** (significant in both, same
direction). Pathways in the first three categories represent genuine
differences between the two training modalities; concordant pathways
are explicitly excluded from that "different" set, since those are
pathways that respond the same way regardless of modality. NOTE: no
significant discordant pathways were identified.

```{r gsea-diff-counts, results='asis'}
include_fig(p$gsea_diff_counts, caption = "Pathway classification counts by status and database, Training_BFR vs Training_HLRT.", width = "75%")
```
